\bf Estimating the number of clusters using Cross Validation

### 2015-12-16 New simulation/technometrics.tex
Add error bar, more consist theory and some editing

Among all the proposed methods, very few provide theoretical consistency result for choosing $k$. 
Assume data is a mixture of $G$ $p$-dimension clusters with equal prior, identically distributed with 
common covariance matrix and finite fourth moments in each dimension, \citet{sugar2003finding} shows
the Jump method will pick the correct $k$ under the conditions that cluster centers are sufficiently
separated and an appropriate transformation is used. \cite{tibshirani2005cluster} proved that the prediction 
strength method is consistent under the setting that the $k$ clusters are uniformly distributed on $k$ unit balls
in $p$-dimension, with pairwise distance of centers are at least four.

@@

\section{Simulation}
In this section, simulation is performed to evaluate the performance of our
proposed methods in locating the ``correct" number of clusters. We compare
with a basket of existing methods including Gap statistics
\citep{tibshirani2001estimating}, Gaussian mixture model-based clustering
\citep{fraley2002model}, CH-index \citep{calinski1974dendrite}, Hartigan
statistics \citep{hartigan1975clustering}, Jump method
\citep{sugar2003finding}, Prediction strength \citep{tibshirani2005cluster},
Bootstrap stability \citep{fang2012selection} in following simulation
settings. We select $p=q$ and $5-$fold cross-validation in row
($m=\frac{1}{4}n$) as default parameter setting for our proposed Gabriel
method. Note that set $p=q$ corresponding to $2-$fold cross-validation in
column.

@@

  \item \textbf{Correlation between dimensions} -- Six clusters in $10$ dimensions. 
  	Each cluster has $100$ or $50$
    multivariate normal observations with common covariance matrix $\Sigma$ 
    which has compound symmetric structure with $1$ in diagonal and $\rho$ 
    off diagonal. For each $\rho$, $100$ data are simulated and the proportion each
    method successfully picks $k=6$ is reported. $\rho$ takes value in $\{0,0.1,...,0.9\}$.

	\begin{figure}[H]
	\centering
	\includegraphics[width=5.5in, height=3.3in]{demo/bench/setting1/Overlay.pdf}
	\label{fig:setting1}
	\end{figure}

  \item \textbf{Noise dimensions} --- Three clusters in $6$ dimensions. Each cluster has $100$ or $50$
    mean zero multivariate normal observations with identity covariance matrix. 
    We add $P$ dimensions of noise to the data, which is randomly generated from uniform$[0,1]$. The noise
    dimension $P$ takes values in $\{0,6,...,54\}$.

	\begin{figure}[H]
	\centering
	\includegraphics[width=5.5in, height=3.3in]{demo/bench/setting2/Overlay.pdf}
	\label{fig:setting2}
	\end{figure}

@@

	\begin{figure}[H]
	\centering
	\includegraphics[width=5.5in, height=3.3in]{demo/bench/setting3/Overlay.pdf}
	\label{fig:setting3}
	\end{figure}

@@

	\begin{figure}[H]
	\centering
	\includegraphics[width=5.5in, height=3.3in]{demo/bench/setting4/Overlay.pdf}
	\label{fig:setting4}
	\end{figure}

	\item \textbf{Heavy tail data} --- Five clusters in $15$ dimensions, each has
     $80$ observations in it. Observations have independent $t$ distribution in each dimension  
     with degree of freedom $\nu$, which takes value in $\{2,3,...,11\}$

	\begin{figure}[H]
	\centering
	\includegraphics[width=5.5in, height=3.3in]{demo/bench/setting5/Overlay.pdf}
	\label{fig:setting4}
	\end{figure}
\end{enumerate}

Note that in all settings above, cluster centers are randomly generated from 
multivariate normal distribution $\mathcal{N}\left(\mathbf{0},\varsigma \mathbf{I}\right)$. 
All clusters are well-separated, i.e. no overlapping. In fact, any simulated 
clusters with minimum distance less than $1$ unit was discarded, so there is
clear definition of true number of clusters. The parameters $\varsigma$ is chosen 
such that about half of the random realization were discarded. The idea is borrowed from
\cite{tibshirani2001estimating}.

@@

\section{Discussion}
\label{sec:conc}
In this paper, we proposed a novel approach to estimate the number of
clusters. The intuition behind our proposed methods is to transfer the
unsupervised learning problem into supervised learning problem via novel form
of cross validation. Such approach is quite different from previous methods which 
utilize the within/between cluster dispersion or stability criterion for selecting the 
optimal $k$. Our method utilizes the connection between different dimensions (columns)
of data through the uniqueness of each cluster center. We proved the self-consistency 
for our proposed Gabriel CV method as well as its asymptotic property with Gaussian noise,
and showed the robustness of our method by simulation. Our method has very good performance
in our limited simulation settings and real data application, and clearly the superior
one when data is high dimension or is heavy-tailed.

Besides no modeling assumption is required, our proposed method is robust against data set 
with variance heterogeneity, unequal number of observations, non-Gaussian noise and high-dimension. 
Such robustness is important in practice because for any given data, it's hard to tell whether
its clusters have different number of observations, is the variance equal for each cluster,
or what underlying noise distribution it has. The weakness of our proposed method is 
that its theory assumes only week correlation between ``predictor'' columns and ``response'' 
columns. In practice, many data sets don't exhibit high correlations between columns, where
our proposed method can be safely applied. In case the high correlation does exist, some procedure
such as rotating the data set and/or leave out most columns for clustering may be used to
reduce its effect as previously mentioned. However, the theory for those procedure have not
been fully developed and it could be the future research topic.

>>>>

Some proposed methods provide theoretical consistency result for choosing $k$. 
Assume data is a mixture of $G$ $p$-dimension clusters with equal prior, identically distributed with 
common covariance matrix and finite fourth moments in each dimension, \citet{sugar2003finding} shows
the Jump method will pick the correct $k$ under the conditions that cluster centers are sufficiently
separated and an appropriate transformation is used. Specifically, let $\Delta\sqrt{p}$ denotes the 
minimum Euclidean distance between cluster centers, \citet{sugar2003finding} shows that the jump will
be maximized at $k=G$ given that $\Delta > 6$ and the existence of a positive $Y$ such that
\[	\left(\frac{p\Delta^2W}{9G}\right)^{-Y}+\left(W\left[\frac{2^{2H^*(X)}}{K^2_{max}2\pi e}\right]-(\frac{\Delta}{6})^2(1-W) \right)^{-Y} < 2	\]
and $\left(\frac{p\Delta^2W}{9G}\right)^{-Y} <1/2$, where $H^*(X)$ is the minimum entropy of the cluster 
membership over each dimension and $W =1- \frac{6^4V_{\mathbf{X}}}{(\Delta^2-36)^2}$ with $V_{\mathbf{X}} = var\left( 1/p ||\mathbf{X}-\mu_j||^2_{\Gamma^{-1}} \mid \mathbf{X} \hspace{0.05in}\text{in jth cluster}\right)$

\cite{tibshirani2005cluster} proved that the prediction strength method is consistent under the setting 
that the observations in $p$-dimension are uniformly distributed within one unit from their respective cluster centers, with minimum distance between the $G$ cluster centers is four unit. Under such well-separated setting, 
they show that 
\[ ps(G) = 1+ \OhP(1), \hspace{0.5in} \underset{k \geq G+1}{\text{sup}} ps(k)\leq \frac{2}{3}+ \OhP(1) \]
therefore $\hat{k}$ is consistent in estimating $G$.  
When data does come from mixture of Gaussian distribution, the model-based method of \citet{fraley2002model} is consistent in estimating $k$ in low dimension.

@@

\section{Simulation}
In this section, simulation is performed to evaluate the performance of our
proposed methods in locating the ``correct" number of clusters. We compare
with a basket of existing methods including Gap statistics
\citep{tibshirani2001estimating}, Gaussian mixture model-based clustering
\citep{fraley2002model}, CH-index \citep{calinski1974dendrite}, Hartigan
statistics \citep{hartigan1975clustering}, Jump method
\citep{sugar2003finding}, Prediction strength \citep{tibshirani2005cluster},
Bootstrap stability \citep{fang2012selection} in following simulation
settings. All methods are executed with their default parameter settings. 
We select $p=q$ and $5-$fold cross-validation in row
($m=\frac{1}{4}n$) as default parameter setting for our proposed Gabriel
method. Note that set $p=q$ corresponding to $2-$fold cross-validation in
column. Wold method (detailed in Appendix $A$) and correlation-corrected 
Gabriel method (section $4.3$) are also included for comparison.

Note that in all settings, cluster centers are randomly generated from 
multivariate normal distribution $\mathcal{N}\left(\mathbf{0},\varsigma \mathbf{I}\right)$. 
All clusters are well-separated, i.e. no overlapping. In fact, any simulated 
clusters with minimum distance less than one unit was discarded, so there is
clear definition of true number of clusters. The parameters $\varsigma$ is chosen 
such that about half of the random realization were discarded. The idea is borrowed from
\cite{tibshirani2001estimating}. The proportion of each method successfully picks the correct
$k$ out of $100$ simulation trials is reported, along with its corresponding confident interval
by Wilson's method \citep{wilson1927probable}.

@@

  \item \textbf{Correlation between dimensions} -- Six clusters in $10$ dimensions. 
  	Each cluster has $100$ or $50$
    multivariate normal observations with common covariance matrix $\Sigma$ 
    which has compound symmetric structure with $1$ in diagonal and $\rho$ 
    off diagonal. $\rho$ takes value in $\{0,0.1,...,0.9\}$.

	\begin{figure}[H]
	\centering
	\includegraphics[width=5.5in, height=3.3in]{demo/bench/setting1/Facet.pdf}
	\label{fig:setting1}
	\end{figure}

  \item \textbf{Noise dimensions} --- Three clusters in $6$ dimensions. Each cluster has $1000$ or $500$
    mean zero multivariate normal observations with identity covariance matrix. 
    We add $P$ dimensions of noise to the data, which is randomly generated from uniform$[0,1]$. The noise
    dimension $P$ takes values in $\{0,6,...,54\}$.

	%\begin{figure}[H]
	%\centering
	%\includegraphics[width=5.5in, height=3.3in]{demo/bench/setting2/Facet.pdf}
	%\label{fig:setting2}
	%\end{figure}

@@

	\begin{figure}[H]
	\centering
	\includegraphics[width=5.5in, height=3.3in]{demo/bench/setting3/Facet.pdf}
	\label{fig:setting3}
	\end{figure}

@@

	\begin{figure}[H]
	\centering
	\includegraphics[width=5.5in, height=3.3in]{demo/bench/setting4/Facet.pdf}
	\label{fig:setting4}
	\end{figure}

	\item \textbf{Heavy tail data} --- Five clusters in $15$ dimensions, each has
     $80$ observations in it. Observations have independent $t$ distribution in each dimension  
     with degree of freedom $\nu$, which takes value in $\{11,10,...,2\}$

	\begin{figure}[H]
	\centering
	\includegraphics[width=5.5in, height=3.3in]{demo/bench/setting5/Facet.pdf}
	\label{fig:setting4}
	\end{figure}
\end{enumerate}

From the graph of setting $1$, we can see high correlation between dimensions cause problem for most 
existing methods as well as the proposed Gabriel CV method and Wold method. Higher correlation clearly 
degenerate most algorithms' ability to find the correct $k$ as data no longer spherical in $p$-dimension.
And the recent proposed methods (Gabriel CV method, Gap statistic, Jump method) seem to be more sensitive to
high correlation than those old methods such as CH-index and Hartigan statistics. The only two methods 
work well in high correlation are the Gaussian model-based BIC method \citep{fraley2002model} and the 
correlation-corrected Gabriel method.

Plot of setting $3$ shows some methods are insensitive to higher dimension such as Jump method and Gap 
statistics, while others deteriorate quickly with increasing dimension, most notably the Gaussian 
model-based BIC method. Note here the data are still generated from mixture of Gaussian. In contrast, our 
proposed Gabriel CV method as well as its correlation-corrected version actually working better in higher 
dimension.

Result of setting $4$ indicates most previously methods are sensitive to variance heterogeneity among 
clusters, most notably the Gap statistics and model-based BIC method. The proposed Gabriel CV method
and its correlation-corrected version consistently perform well in estimating $k$ and quite insensitive
to variance heterogeneity. The Wold method also performs well in this setting.

The last non-Gaussian setting examines methods' performance in heavy-tail data. With degree of freedom 
decreases, the tail becomes more flat and the Gaussian assumption becomes more inappropriate. It explains
the sharp plunge of Gaussian model-based BIC method in the plot. For most methods, their performances are 
relatively stable until the tail gets very heavy. In case where degree of freedom is $2$, Gap statistics 
and Jump method dived considerably compare to the Gabriel method and Wold method.

In conclusion, the proposed Gabriel CV method compares well with current state-of-the-art methods and 
is very robust against variance heterogeneity, high dimension and heavy-tail data. One weakness of the 
Gabriel CV method is that it's sensitive to the high correlation among dimensions. In such case, its 
correlation-corrected version will perform well under the assumption of common covariance structure.

@@

\section{Discussion}
\label{sec:conc}
In this paper, we proposed a novel approach to estimate the number of
clusters. The intuition behind our proposed methods is to transfer the
unsupervised learning problem into supervised learning problem via novel form
of cross validation. Such approach is quite different from previous methods which 
utilize the within/between cluster dispersion or stability criterion for selecting the 
optimal $k$. Our method utilizes the connection between different dimensions (columns)
of data through the uniqueness of each cluster center. We proved the self-consistency 
for our proposed Gabriel CV method as well as its asymptotic property with Gaussian noise,
and showed the robustness of our method by simulation. Our method has very good performance
in our limited simulation settings and real data application, and clearly the superior
one when data has heterogeneous variance or is heavy-tailed.

Besides no strong modeling assumption is required, our proposed method is robust against data set 
with variance heterogeneity, unequal number of observations, non-Gaussian noise and high-dimension. 
Such robustness is important in practice because for any given data, it's hard to tell whether
its clusters have different number of observations, is the variance equal for each cluster,
or what underlying noise distribution it has. The weakness of our proposed method is 
that its theory assumes only week correlation between ``predictor'' columns and ``response'' 
columns. In practice, many data sets don't exhibit high correlations between columns, where
our proposed method can be safely applied. In case the high correlation does exist, 
the correlation-corrected version of Gabriel CV method in section $4.3$ can be used if we can 
assume common covariance structure. Other procedure such as leave out most columns for clustering 
may also be used to reduce its effect. However, the theory for such procedure has not
been fully developed and it could be the future research topic.
